Reducing Acceptance Marks in Emerson-Lei Automata by QBF-solving

### 2022-04-26 paper/main.tex
final polishing

\begin{table}[t]
\caption{The cummulative numbers of acceptance marks before and after reduction
for various reductions and automata sets.
%\textit{X/Y solved aut.} means that out of
%\textit{Y} successfully translated LTL formulas, every run of \telatko ended
%within 120 seconds time limit on \textit{X} automata. We sum up the acceptance
%marks only on these automata.
The column \textit{solver timeout} indicates the number of instances
when the last call to the QBF solver timed out.  }
\label{tab:cumm-data}
\centering
\setlength{\tabcolsep}{10pt}
\input{table-data.tex}
\end{table}

Table~\ref{tab:heatmap} shows the number of automata coming from all
four sets that have a given number of original acceptance marks and a
given number of acceptance marks after multi-level reduction. The
table indicates that in many cases only 1 or 2 marks can be
saved. However, the achieved reduction is substantial for some
automata with a higher number of original acceptance marks. For
example, in 24 cases, we have successfully reduced 7 or more
acceptance marks to only 4.
% Moreover, there were many timeouts of the last QBF query for manof these automata.
% Let us note that some that some reductions are the results are promising, e.g. in 17 cases, we have successfully
% reduced 8 or more acceptance marks to only 4.

\begin{table}[t]
\caption{The effect of reduction.
A cell on coordinates $(x,y)$ contains the number of automata that have been
reduced from $x$ to $y$ acceptance marks. If the cell contains a sum of two
numbers, the latter represents number of automata where the last query to QBF
solver has not been answered in the time limit.}
\label{tab:heatmap}
\setlength{\tabcolsep}{4pt}
\centering
\input{table-heatmap.tex}
\end{table}

@@

We have presented an application of QBF solving to reduction of the number of
acceptance marks in transition-based Emerson-Lei automata. The method
has been implemented and our experimental results show that it is possible
to use the QBF solvers to find equivalent acceptance conditions of TELA
with fewer acceptance marks and without changes to automata structure.

The reduction of acceptance marks is not the only application of the
presented approach: one could possibly look for, e.g., equivalent
parity automata or automata without $\Fin$-marks, depending on the
subsequent use of the automata. Even though the QBF queries can be
time-consuming, in practice it can be well used and one can find a
good trade-off between speed and efficiency by choosing the precision of
the reduction and setting a reasonable timeout.

In the future work, it could be examined how the reductions of the acceptance
condition can contribute to further reduction of the state space by enabling
various optimizations, e.g., simulations.

>>>>

\begin{table}[t]
  \caption{The cummulative numbers of acceptance marks before and
    after reduction for various reductions and automata sets. The
    column \emph{reduction} shows the cummulative percentage of saved
    acceptance marks and \emph{time} reports the cummulative reduction
    time.
%\textit{X/Y solved aut.} means that out of
%\textit{Y} successfully translated LTL formulas, every run of \telatko ended
%within 120 seconds time limit on \textit{X} automata. We sum up the acceptance
%marks only on these automata.
The column \textit{solver timeout} indicates the number of instances
when the last call to the QBF solver timed out.  }
\label{tab:cumm-data}
\centering
\setlength{\tabcolsep}{10pt}
\input{table-data.tex}
\end{table}

Table~\ref{tab:heatmap} shows the number of automata from all
four sets that have a given number of original acceptance marks and a
given number of acceptance marks after multi-level reduction. The
table indicates that in many cases only 1 or 2 marks can be
saved. However, the achieved reduction is substantial for some
automata with a higher number of original acceptance marks. For
example, in 24 cases, we have reduced 7 or more
acceptance marks to only 4.
% Moreover, there were many timeouts of the last QBF query for manof these automata.
% Let us note that some that some reductions are the results are promising, e.g. in 17 cases, we have successfully
% reduced 8 or more acceptance marks to only 4.

\begin{table}[t]
\caption{The effect of reduction.
A cell on coordinates $(x,y)$ contains the number of automata that have been
reduced from $x$ to $y$ acceptance marks. If the cell contains a sum of two
numbers, the latter represents number of automata where the last call to QBF
solver reached the time limit.}
\label{tab:heatmap}
\setlength{\tabcolsep}{4pt}
\centering
\input{table-heatmap.tex}
\end{table}

@@

We have presented a method reducing the number of acceptance marks in
transition-based Emerson-Lei automata with use of QBF solving and
without altering automata structure. We have implemented the method in
a tool called \texttt{telatko}. The presented experimental results
show that the tool can indeed reduce the number of acceptance marks in
automata produced by state-of-the-art LTL to automata translators. The
reduction was significant on automata produced by the translators of
the Owl library.

The reduction of acceptance marks is not the only application of the
presented approach. For example, it can be easily adopted to look for
an equivalent automaton with the same structure and an acceptance
formula of a specific form (e.g., without any $\Fin m$ terms). Even
though the QBF queries can be time-consuming, in practice one can
often find a good trade-off between speed and efficiency by adjusting 
the formula precision and choosing a reasonable timeout.

In future, we plan to improve the reduction process (e.g., use the
bisection method to reduce the number of acceptance marks more
efficiently) and study whether the reduction of acceptance condition
can enable further reductions or simplifications of automata
structures.

### 2022-04-26 paper/table-data.tex
final polishing

\midrule
\multirow{4}{*}{\makecell{\texttt{ltl2dela}\\129 automata\\523 acc.~marks}}
&
level 1 &
387 & 26.0 & 755.4 &  18
\\
&
level 2 &
390 & 25.4 & 1024.5 &  20
\\
&
level 3 &
393 & 24.9 & 7273.3 &  25
\\
&
multi &
371 & 29.1 & 9111.4 &  21
\\
\midrule
\multirow{4}{*}{\makecell{\texttt{ltl2dgra}\\234 automata\\882 acc.~marks}}
&
level 1 &
552 & 37.4 & 738.7 & 14
\\
&
level 2 &
569 & 35.5 & 907.3 &  18
\\
&
level 3 &
561 & 36.4 & 1163.4 &  21
\\
&
multi &
542 & 38.5 & 2320.7 &  21
\\
\midrule
\multirow{4}{*}{\makecell{\texttt{ltl2tgba}\\70 automata\\198 acc.~marks}}
&
level 1 &
198 & 0.0 & 49.3 &  0
\\
&
level 2 &
198 & 0.0 & 78.9 & 0
\\
&
level 3 &
189 & 4.5 & 1222.8 & 7
\\
&
multi &
189 & 4.5 & 1259.1 & 6
\\
\midrule
\multirow{4}{*}{\makecell{\texttt{ltl3tela}\\91 automata\\348 acc.~marks}}
&
level 1 &
332 & 4.6 & 465.9 &  12
\\
&
level 2 &
334 & 4.0 & 553.9 &  15
\\
&
level 3 &
326 & 6.3 & 691.8 &  18
\\
&
multi &
323 & 7.2 & 1545.9 & 18
\\

>>>>

\midrule
\multirow{4}{*}{\makecell{\texttt{ltl2dela}\\129 automata\\523 acc.~marks}}
&
level 1 &
387~ & 26.0~~~~~~ & 755.4~ &  18~~~
\\
&
level 2 &
390~ & 25.4~~~~~~ & 1024.5~ &  20~~~
\\
&
level 3 &
393~ & 24.9~~~~~~ & 7273.3~ &  25~~~
\\
&
multi &
371~ & 29.1~~~~~~ & 9111.4~ &  21~~~
\\
\midrule
\multirow{4}{*}{\makecell{\texttt{ltl2dgra}\\234 automata\\882 acc.~marks}}
&
level 1 &
552~ & 37.4~~~~~~ & 738.7~ & 14~~~
\\
&
level 2 &
569~ & 35.5~~~~~~ & 907.3~ &  18~~~
\\
&
level 3 &
561~ & 36.4~~~~~~ & 1163.4~ &  21~~~
\\
&
multi &
542~ & 38.5~~~~~~ & 2320.7~ &  21~~~
\\
\midrule
\multirow{4}{*}{\makecell{\texttt{ltl2tgba}\\70 automata\\198 acc.~marks}}
&
level 1 &
198~ & 0.0~~~~~~ & 49.3~ &  0~~~
\\
&
level 2 &
198~ & 0.0~~~~~~ & 78.9~ & 0~~~
\\
&
level 3 &
189~ & 4.5~~~~~~ & 1222.8~ & 7~~~
\\
&
multi &
189~ & 4.5~~~~~~ & 1259.1~ & 6~~~
\\
\midrule
\multirow{4}{*}{\makecell{\texttt{ltl3tela}\\91 automata\\348 acc.~marks}}
&
level 1 &
332~ & 4.6~~~~~~ & 465.9~ &  12~~~
\\
&
level 2 &
334~ & 4.0~~~~~~ & 553.9~ &  15~~~
\\
&
level 3 &
326~ & 6.3~~~~~~ & 691.8~ &  18~~~
\\
&
multi &
323~ & 7.2~~~~~~ & 1545.9~ & 18~~~
\\
